---
title: "Salmonid Mortality Analysis: TidyTuesday"
subtitle: "Reproducible analysis using quarto"
author: "Arush Mohan"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true               # Table of contents
    toc-depth: 2
    number-sections: true
    code-fold: true         # Allows folding of long code blocks
    embed-resources: true   # Makes a standalone HTML (easy to share)
editor: visual
---

``` {r}
#| label: setup
#| include: false

# Loading in the packages

library(tidyverse)    
library(lubridate)    
library(janitor)      
library(scales)       
library(gt)     
# Setting a theme

theme_set(theme_minimal(base_size = 12))

# Loading in the csv files from Github

monthly_losses_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-03-17/monthly_losses_data.csv')
monthly_mortality_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-03-17/monthly_mortality_data.csv')

```


```{r}
#|label: Cleaning-the-data
#|include: false
#|results: hide

# The data provided is already in a clean format

# Quick check of the data structure

glimpse(monthly_losses_data)
glimpse(monthly_mortality_data)


```

## Overview

Add an overview of the project



## How does monthly mortality differ over time? 

A summary of mortality statistics

```{r}
#| label: mortality-over-time-data
#| include: false


# Convert date to Date class and summarise metrics over time
mort_time <- monthly_mortality_data |>
  mutate(date = as.Date(date)) |>
  group_by(species, date) |>
  summarise(
    median_mortality = median(median, na.rm = TRUE),
    q1 = median(q1, na.rm = TRUE),
    q3 = median(q3, na.rm = TRUE),
    .groups = "drop"
  )
glimpse(mort_time)
```

```{r}
#| label: mortality-over-time-plot

# Line plot showing mortality over time by species.
ggplot(mort_time, aes(x = date, y = median_mortality, colour = species)) +
  geom_line(linewidth = 1.5) +
  labs(
    title = "Median Monthly Mortality Over Time",
    subtitle = "Aggregated across all regions",
    x = NULL,
    y = "Median mortality"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme(legend.position = "bottom")

```


## Which region has the lowest mortality? 

Average mortality for each region - overall lowest 15

```{r}
#| label: regional-mortality-data
#| echo: true

# The dataset does not contain raw individual‑fish mortality measurements, only regional monthly medians

# Average median mortality by region
#Since the dataset contains country, county, and area level data, I will filter to include only counties

mort_by_region <- monthly_mortality_data |> 
  mutate(date = as.Date(date)) |> 
  filter(geo_group == "county") |> 
  group_by(region, species) |> 
  summarise(
    avg_median_mortality = mean(median, na.rm = TRUE), #avg of medians
    p10 = quantile(median, 0.10, na.rm = TRUE),
    p90 = quantile(median, .90, na.rm = TRUE),
    n_months = sum(!is.na(median)), #count of months
    .groups = "drop"
  )

# Select the 15 regions with the lowest average mortality.
lowest_regions <- mort_by_region |>
  arrange(avg_median_mortality) |>
  slice_head(n = 15)

lowest_regions
```


Plotting mortality by region

```{r}
#| label: lowest-mortality-plot
#| echo: true

# Filtering for just salmon
#salmon_lowest_region <- lowest_regions |> 
#  filter(species == "salmon")

# Filtering for just rainbow trout
#trout_lowest_region <- lowest_regions |> 
#  filter(species == "rainbowtrout")

# Salmon lowest mortality graph
ggplot(lowest_regions,
       aes(x = reorder(region, avg_median_mortality),
           y = avg_median_mortality,
           fill = species)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Counties with Lowest Average Median Mortality",
    subtitle = "Filtered to geo_group = 'County'",
    x = NULL,
    y = "Average median mortality"
  ) +
  theme(legend.position = "bottom")

```

## What other types of losses are significant?

The losses dataset includes:

- dead
- discarded
- escaped
- other

We reshape the dataset to explore proportions.

```{r}
#|label: loss-long-data
#|include: false

# Convert wide columns (dead, discarded, escaped, other)
# into a long tidy format: loss_type, count.
loss_long <- monthly_losses_data |>
  mutate(date = as.Date(date)) |>
  select(species, date, geo_group, region, losses, dead, discarded, escaped, other) |>
  pivot_longer(
    cols = c(dead, discarded, escaped, other),
    names_to = "loss_type",
    values_to = "count"
  ) |>
  mutate(share = count / losses)   # compute proportion of total loss

loss_long

```

Summarise total losses by type
```{r}
#| label: loss-summary

# For each loss type, compute total and mean share.
loss_summary_table<- loss_long |>
  group_by(loss_type) |>
  summarise(
    total_count = sum(count, na.rm = TRUE),
    avg_share = mean(share, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_count))

loss_summary_table <- loss_summary_table %>%
  rename(
    "Loss Type" = loss_type,
    "Total Count" = total_count,
    "Average Share" = avg_share
  )

# Step 3 — produce a styled GT table
loss_summary_table %>%
  gt() %>%
  
  # Add a title and subtitle
  tab_header(
    title = "Summary of Loss Types",
    subtitle = "Total losses and their average proportion"
  ) %>%
  
  # Format numbers nicely
  fmt_number(
    columns = "Total Count",
    sep_mark = ",", 
    decimals = 0
  ) %>%
  fmt_percent(
    columns = "Average Share",
    decimals = 2
  ) %>%

  # Highlight rows (example: highlight the largest total count)
  data_color(
    columns = "Total Count",
    colors = scales::col_numeric(
      palette = c("#D9EAF7", "#004F9F"),  # light blue → NHS blue
      domain = NULL
    )
  ) %>%

  # Add borders around table + row separators
  tab_style(
    style = cell_borders(
      sides = "all",
      color = "grey60",
      weight = px(1)
    ),
    locations = cells_body()
  ) %>%
  tab_style(
    style = cell_borders(
      sides = "bottom",
      color = "black",
      weight = px(2)
    ),
    locations = cells_column_labels()
  ) %>%

  # Style column headers (remove underscores, nicer background)
  tab_style(
    style = list(
      cell_fill(color = "#E5F0FA"),       # NHS light blue background
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  )
```

Loss over time

```{r}
#| label: loss-over-time

# Summarise total losses by type per month.
loss_time <- loss_long |>
  group_by(loss_type, date) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop")

ggplot(loss_time,
       aes(x = date, y = total, colour = loss_type)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Loss Types Over Time",
    subtitle = "Aggregated across all regions and species",
    x = NULL,
    y = "Number of fish"
  ) +
  scale_y_continuous(labels = comma) +  # prettier y-axis
  theme(legend.position = "bottom")

```

Loss composition by region

```{r}
#| label: loss-by-region-plot

# Compute regional loss composition (share of each type)
loss_by_region <- loss_long |>
  filter(geo_group == "county") |> 
  group_by(region, loss_type) |>
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") |>
  group_by(region) |>
  mutate(share = total / sum(total)) |>
  ungroup()

# Select the 5 regions with the highest total losses.
top_regions <- loss_by_region |>
  group_by(region) |>
  summarise(total_losses = sum(total), .groups = "drop") |>
  arrange(desc(total_losses)) |>
  slice_head(n = 5) |>
  pull(region)

# Plot the composition.
loss_by_region |>
  filter(region %in% top_regions) |>
  ggplot(aes(x = reorder(region, share), y = share, fill = loss_type)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Loss Composition by Region",
    subtitle = "Top 5 counties by total losses",
    x = NULL,
    y = "Share of total losses"
  ) +
  scale_y_continuous(labels = percent_format()) +
  theme(legend.position = "bottom")

```

